Counts of Crimes and Their Subtypes in Toronto from 2014 to 2023

### 2024-09-27 paper/paper.qmd
Fifth Commit. Sketches yet to be uploaded.

---
title: "Counts of Crimes and Their Subtypes in Toronto from 2014 to 2023"
subtitle: "Details and Trends"
author: 
  - Justin (Jiazhou) Bi
thanks: "Inspired and instructed by: https://github.com/RohanAlexander/marriage."
date: Sept. 23, 2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
jupyter: python3 
number-sections: true
bibliography: references.bib
---

@@

Crime is defined as "an action or omission which constitutes an offence and is punishable by law." (@TheOxfordDictionary_1992) However, this seemingly straight forward definition can misguide us to think that crime is simple and clear concept with a specific defining criteria, swift process of judgement, and pre-defined punishment. Crime, in fact, is a complext concept that affects every aspects of all individuals, communities, and the societies. Crimes can reshape the power dynamics of our social structure, influence our public policies, and even change our day-to-day interaction with others. As a result, understanding the trends in crimes and their variations in their subtypes is crucial for policymakers, law enforcements, and every member of our society to devise better strategies to prevent crimes from happening. Therefore, this report aims to explore and examine the current of reported crimes in Toronto from 2014 to 2023, discover and highlight some of fluctuations, as well as propose some potential underlying factors that might have contributed to these trends. The raw data being used in the report is available from Toronto Open Data. (@TorontoOpenData) For the data analysis, I have used pandas (@pandas). For data visualization, I have used seaborn. (@seaborn) All the tables were generated by using plotly. (@plotly)

The analysis of crime data offers valuable insights into not only the frequency of criminal activity but also the shifting nature of these offenses. Some types of crime, such as violent offenses, theft, and property crimes, have historically shown consistent trends, often influenced by broader socio-economic conditions. In contrast, emerging forms of criminal activity, such as cybercrime and fraud, have gained prominence in recent years, reflecting the growing reliance on technology and the internet in everyday life. Understanding these patterns requires a detailed examination of various factors including demographic shifts, law enforcement practices, economic conditions, and legislative changes that may have influenced the reporting and recording of crimes during the studied period.

Between 2014 and 2023, the world experienced significant social, economic, and technological transformations, each of which likely had an impact on crime rates. For instance, the COVID-19 pandemic, which began in 2020, not only disrupted daily routines but also led to shifts in criminal behavior. Lockdowns and other public health measures reduced opportunities for certain types of crime, such as burglaries and public disturbances, while simultaneously leading to increases in domestic violence and cybercrime. Additionally, economic hardship caused by the pandemic, including rising unemployment rates and financial instability, may have driven a surge in financially motivated crimes, such as fraud and theft.

Beyond the pandemic, advancements in technology during this period contributed to the evolution of criminal activity, particularly in the realm of cybercrime. As digital technologies became more integrated into personal and professional life, the potential for cyber-related offenses, including hacking, identity theft, and online fraud, increased exponentially. The digital landscape has also made it more difficult for law enforcement agencies to detect and prevent certain types of crime, as criminals now have the tools to operate anonymously across borders and evade traditional methods of investigation.

Another important consideration in crime trends is the impact of law enforcement practices and legal reforms. Changes in policing strategies, sentencing laws, and public attitudes toward certain offenses can significantly affect the reporting and prosecution of crimes. For instance, the decriminalization of certain substances, such as cannabis, in various regions during this period may have contributed to declines in drug-related arrests, while increased awareness and reporting of issues like sexual assault and domestic violence have brought these crimes into sharper focus in recent years.

In this report, we will delve into crime statistics from 2014 to 2023, breaking down reported crimes into key subtypes such as violent crime, property crime, and cybercrime, among others. By analyzing these trends, we aim to provide a clearer understanding of how crime has evolved over the last decade and identify possible socio-economic, technological, and legal factors that have driven these changes. Through this analysis, we can gain valuable insights into how best to address the challenges posed by both traditional and emerging forms of crime in the future.

The analysis of crime data offers valuable insights into not only the frequency of criminal activity but also the shifting nature of these offenses. Some types of crime, such as violent offenses, theft, and property crimes, have historically shown consistent trends, often influenced by broader socio-economic conditions. In contrast, emerging forms of criminal activity, such as cybercrime and fraud, have gained prominence in recent years, reflecting the growing reliance on technology and the internet in everyday life. Understanding these patterns requires a detailed examination of various factors including demographic shifts, law enforcement practices, economic conditions, and legislative changes that may have influenced the reporting and recording of crimes during the studied period.

The remainder of this paper is structured as follows.

@@

The dataset being used here is downloaded from Toronto Open Data website via @TorontoOpenData. This dataset contains all the reported crimes happened in the city of Toronto from 2014 to 2023. This dataset is grouped by the year of the reported crime, its category and belonging subtype, as well as the count of the subtype being reported and cleared for that year, for each division. Because I am examining the crime pattern in the whole city, I have dropped the division information, and aggregated the existing data according to their subtype, and the year of the crime being reported. In the following sub-sections, I will go through all the variables used in this report, and provide some basic descriptive statistics about them. The first 5 rows of the cleaned data used for analysis is attached (@tbl-table1).

@@

## Report Year
Report Year variable is the year of crimes being reported. In this dataset, the data spans from 2014 to 2023, encompassing a ten-year period. No month or date information was given, thus there are only 10 different values for this variable, in a chronological order.

## Category
Category includes information about the nature of the crime. In total there are six crime categroies, they are Crimes Against Property, Crimes Against the Person, Other Federal Statute Violations, Other Criminal Code Violations, Controlled Drugs and Substances Act, and Criminal Code Traffic. They are listed in the table below (@tbl-table2).

@@

## Subtype
Under each crime category, there exist multiple subtypes within each category. The following is an exhaustive table (@tbl-table3) of all of the subtypes of crimes and their respective category.

@@

## Count
In the original table, this value is grouped by the subtype of the crime, the division, and the year when the crime was reported. The original count indicates the number of a certain subtype of crime reported within a certain division for the year. However, as mentioned before, because I am only interested in looking at all the crimes for the City of Toronto in general, I have dropped the division information, and aggregated the count from all the divisions to a single value. Therefore, for each subtype of the crimes, there is a total count of that subtype being reported in a single year.

## Count_Cleared
This is the counts of crimes identified as cleared. In plain words, these are crimes that are dealt/solved. I have taken the same approach for this column as the previous one. After cleaerning the data,f or each subtype of the crimes, there is a total count of that subtype being reported that is also cleaered in a single year.

## Case_Clearing_Rate
This column was not included in the raw dataset, but created by dividing the cleared crimes by total crimes. A higher case clearing rate for a particulat subtype of crime usually suggests higher effectiveness of law enforcements in dealing with this subtype of crime. The value is ranged from 0 to 100%.

# Results
## Trend of Reported Cases by Year
During this ten-year-period (2014 - 2023), the overall counts of cases being reported in Toronto has increased significantly, from 113,426 in 2014 to 169,620 in 2023. Meanwhile, a noticable decrease of reported crime counts dropped in 2020 (118,568 cases) and 2021 (120,397 cases), compared to 2019 (144,529 cases). @tbl-table4 below shows the aggregated counts of reported crimes, grouped by their reported years.

@@

The @fig-figure1 belows depicts the changing pattern of the amounts of reported crimes by year.

@@

## Trend of Reported Cases per Category by Year
Among all the six categories, crimes against property has a significant higher count than all the other categories. In 2023, a total of 119,629 cases of crimes against property were reported, compared to the second highest count, which is crimes against ther person, with 29,892 cases. Below is a table of the count of reported crime for each category, for 2014, 2020, and 2023, as well as a line chart for the flucuating trend. The @tbl-table5 below shows the amount of reported cases for each category for 2014, 2020, and 2023, respectively.

@@

The following is @fig-figure2 showing the trend of how counts of reported crimes varied from 2014 to 2023 for each category.

@@

## Subtype of Crimes Against Property by Year and Their Gorwth Rates
The overall pattern for each subtype of crimes under crimes against property is comparable to the overall reported crime pattern, with a increase from 2014 to 2023, despite a significant drop in 2020 and 2021. Among all the subtypes within crimes against property, the most common reported crime is theft under $5,000. For each subtype, the nominal growth rate between 2014 and 2023 is also calculated and attached here.
```{python}
#| echo: false
#| warning: false
#| label: tbl-table6
#| tbl-cap: Growth Rate of Each Crime Against Property Subtype from 2014 to 2023
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')

@@

@fig-figure3 here shows the amount of reported cases for each subtype of crime against property for 2014 and 2023, side by side, respectively.

@@

## Crime Subtype Clear Rate from 2014 to 2023 and Its Trend
Crime clearing rate has dropped for all subtypes of crimes against property from 2014 to 2023, ranging from a 10.25% decrease for break and enter to apartment, to 87.02% decrease for fraud. The clearing rate for 2014 and 2023, as well as the change in clear rate for each sutype of crime against property are listed below in @tbl-table7.

```{python}
#| echo: false
#| warning: false
#| label: tbl-table7
#| fig-cap: Case Clear Rate and Change from 2014 to 2023 for Crime Against Property Subtypes
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')
crime_against_property_df = df[df['Category'] == 'Crimes Against Property']
grouped_by_year_subtype_df = crime_against_property_df.groupby(['Year', 'Subtype'])['ClearRate'].mean().reset_index()
pivoted_clear_rate_df = grouped_by_year_subtype_df.pivot(index='Subtype', columns='Year', values='ClearRate')
pivoted_clear_rate_df = pivoted_clear_rate_df[[2014, 2023]]
pivoted_clear_rate_df['Clear Rate Change'] = (((pivoted_clear_rate_df[2023]-pivoted_clear_rate_df[2014])/pivoted_clear_rate_df[2014])*100).round(2)
pivoted_clear_rate_df['Clear Rate Change']=pivoted_clear_rate_df['Clear Rate Change'].astype(str)+'%'
pivoted_clear_rate_df.columns.name = None
pivoted_clear_rate_df.reset_index(inplace=True)
pivoted_clear_rate_df.index=pivoted_clear_rate_df.index +1
pivoted_clear_rate_df

@@

@fig-figure4 show the trend of the change in case clearing rate of each subtype of crime against property from 2014 to 2023.

@@

# Discussion

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "Counts of Crimes and Their Subtypes in Toronto from 2014 to 2023"
subtitle: "Details and Trends"
author: 
  - Justin (Jiazhou) Bi
thanks: "Inspired and instructed by: https://github.com/RohanAlexander/marriage."
date: Sept. 23, 2024
date-format: long
abstract: "This report analyzes crime trends and subtypes in Toronto from 2014 to 2023, highlighting fluctuations and speculating possible socio-economic, technological, and pandemic-related factors influencing criminal activity. Using data from Toronto Open Data, the study aggregates and examines reported crimes in categories such as property crimes, violent offenses, and cybercrime. The results show a significant increase in reported crimes, from 113,426 in 2014 to 169,620 in 2023. A notable drop in crime during 2020 and 2021 corresponds with COVID-19 lockdowns, which reduced certain physical crimes while increasing cybercrimes like fraud. Despite the overall rise in crime, clearance rates for crimes like fraud and theft significantly declined, underscoring challenges for law enforcement."
format: pdf
jupyter: python3 
number-sections: true
bibliography: references.bib
---

@@

Crime is defined as "an action or omission which constitutes an offense and is punishable by law." (@TheOxfordDictionary_1992) However, this seemingly straightforward definition can misguide us to think that crime is a simple and clear concept with specific defining criteria, a swift process of judgment, and pre-defined punishment. Crime is a complete concept that affects every aspect of all individuals, communities, and societies. Crimes can reshape the power dynamics of our social structure, influence our public policies, and even change our day-to-day interactions with others. As a result, understanding the trends in crimes and the variations in their subtypes is crucial for policymakers, law enforcement, and every member of our society to devise better strategies to prevent crimes from happening. Therefore, this report aims to explore and examine the current of reported crimes in Toronto from 2014 to 2023, discover and highlight some fluctuations, and propose some potential underlying factors that might have contributed to these trends. The raw data being used in the report is available from Toronto Open Data. (@TorontoOpenData) For the data analysis, I have used pandas (@pandas). For data visualization, I have used Seaborn. (@seaborn) All the tables were generated by using Plotly. (@plotly)

This analysis can be helpful by providing valuable insights into current crime activity trends. Certain crimes, such as violent offenses, theft, and property crimes, are believed to be highly correlated with broader socio-economic conditions. (@lofstrom2016crime) Meanwhile, with new technology emerging, other types of criminal activity, such as technology-facilitated fraud, are becoming more prominent. According to RCMP's report, Canadian victims have lost 530 million dollars due to fraud. Thus, examining various types of crime can assist us to better understand these crimes and devise better strategies to better protect our safety and properties.

On the other hand, between 2014 and 2023, the world experienced significant transformations that may impact crime rates. In particular, the COVID-19 pandemic, which began in 2020 and ended in early 2022, has disrupted our daily routines and led to shifts in criminal behaviour. Lockdowns and other public health measures reduced opportunities for certain types of crime, such as burglaries and public disturbances, while simultaneously leading to increases in domestic violence and cybercrime. Additionally, economic hardship caused by the pandemic, including rising unemployment rates and financial instability, may have driven a surge in financially motivated crimes, such as fraud and theft.

This report will delve into crime statistics from 2014 to 2023 in Toronto. I will closely examine all the reported crime categories and specific subtypes of crimes. By analyzing these trends, I aim to clarify how crime has evolved over the last decade and identify possible socio-economic, technological, and legal factors that have driven these changes. Through this analysis, we can gain valuable insights into how best to address the challenges posed by both traditional and emerging forms of crime in the future.

@@

The dataset used here is downloaded from the Toronto Open Data website via @TorontoOpenData. This dataset contains all the reported crimes that happened in Toronto from 2014 to 2023. This dataset is grouped by the year of the reported crime, its category and belonging subtype, and the count of the subtype being reported and cleared for that year for each division. Because I am examining the crime pattern in the city, I have dropped the division information and aggregated the existing data according to their subtype and the year of the crime being reported. In the following subsections, I will review all the variables used in this report and provide some basic descriptive statistics. The first five rows of the cleaned data used for analysis are attached (@tbl-table1).

@@

## Report Year
The report year variable is the number of crimes being reported. In this dataset, the data spans from 2014 to 2023, encompassing ten years. No month or date information was given; thus, there are only ten different values for this variable in chronological order.

## Category
The category includes information about the nature of the crime. There are six crime categories: Crimes Against Property, Crimes Against the Person, Other Federal Statute Violations, Other Criminal Code Violations, Controlled Drugs and Substances Act, and Criminal Code Traffic. They are listed in the table below (@tbl-table2).

@@

## Subtype
There exist multiple subtypes under each crime category. The following is an exhaustive table (@tbl-table3) of all crimes' subtypes and their respective category.

@@

## Count
In the original table, this value is grouped by the subtype of the crime, the division, and the year when the crime was reported. The original count indicates the number of a specific subtype of crime reported within a particular division for the year. However, as mentioned before, because I am only interested in all the crimes in the City of Toronto, I have dropped the division information and aggregated the count from all the divisions to a single value. Therefore, for each subtype of the crimes, a total count of that subtype is reported in a single year.

## Count_Cleared
These are the counts of crimes identified as cleared. In plain words, these are crimes that are dealt/solved. I have taken the same approach for this column as the previous one. After cleaning the data,f, or each subtype of the crimes, there is a total count of that subtype being reported that is also cleared in a single year.

## Case_Clearing_Rate
This column was not included in the raw dataset but was created by dividing the cleared crimes by total crimes. A higher case-clearing rate for a particular subtype of crime usually suggests a higher effectiveness of law enforcement in dealing with this subtype of crime. The value is ranged from 0 to 100%.

# Results
## Trend of Reported Cases by Year
During this ten-year-period (2014 - 2023), the overall cases reported in Toronto have increased significantly, from 113,426 in 2014 to 169,620 in 2023. Meanwhile, a noticeable decrease in reported crime counts dropped in 2020 (118,568 cases) and 2021 (120,397 cases) compared to 2019 (144,529 cases). @tbl-table4 below shows the aggregated counts of reported crimes, grouped by their reported years.

@@

The @fig-figure1 below depicts the changing pattern of reported crimes per year.

@@

## Trend of Reported Cases per Category by Year
Among all the six categories, crimes against property have a significantly higher count than all the other categories. In 2023, 119,629 crimes against property were reported, compared to the second highest count, crimes against the person, with 29,892 cases. Below is a table of the count of reported crimes for each category for 2014, 2020, and 2023, as well as a line chart for the fluctuating trend. The @tbl-table5 below shows the reported cases for each category for 2014, 2020, and 2023, respectively.

@@

The following is @fig-figure2, which shows how the counts of reported crimes varied for each category from 2014 to 2023.

@@

## Subtype of Crimes Against Property by Year and Their Gorwth Rates
The overall pattern for each subtype of crimes under crimes against property is comparable to the overall reported crime pattern, with an increase from 2014 to 2023, despite a significant drop in 2020 and 2021. Among all the subtypes within crimes against property, the most commonly reported crime is theft under $5,000. The nominal growth rate between 2014 and 2023 for each subtype is also calculated and attached here in @tbl-table6.

```{python}
#| echo: false
#| warning: false
#| label: tbl-table6
#| tbl-cap: Growth Rate of Each Crime Against Property Subtype from 2014 to 2023
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')

@@

@fig-figure3 here shows the number of reported cases for each subtype of crime against property for 2014 and 2023, side by side, respectively.

@@

## Crime Subtype Clear Rate from 2014 to 2023 and Its Trend
The crime-clearing rate has dropped for all subtypes of crimes against the property from 2014 to 2023, ranging from a 10.25% decrease for a break and entry to the apartment to an 87.02% decrease for fraud. The clearing rate for 2014 and 2023 and the change in clear rate for each subtype of crime against property are listed below in @tbl-table7.

```{python}
#| echo: false
#| warning: false
#| label: tbl-table7
#| tbl-cap: Case Clear Rate and Change from 2014 to 2023 for Crime Against Property Subtypes
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')
crime_against_property_df = df[df['Category'] == 'Crimes Against Property']
grouped_by_year_subtype_df = crime_against_property_df.groupby(['Year', 'Subtype'])['ClearRate'].mean().reset_index()
pivoted_clear_rate_df = grouped_by_year_subtype_df.pivot(index='Subtype', columns='Year', values='ClearRate')
pivoted_clear_rate_df = pivoted_clear_rate_df[[2014, 2023]]
pivoted_clear_rate_df['Clear Rate Change'] = (((pivoted_clear_rate_df[2023]-pivoted_clear_rate_df[2014])/pivoted_clear_rate_df[2014])*100).round(2)
pivoted_clear_rate_df['Clear Rate Change']=pivoted_clear_rate_df['Clear Rate Change'].astype(str)+'%'
pivoted_clear_rate_df.columns.name = None
pivoted_clear_rate_df.reset_index(inplace=True)
pivoted_clear_rate_df.index=pivoted_clear_rate_df.index +1
pivoted_clear_rate_df

@@

@fig-figure4 shows the trend of the change in case clearing rate of each subtype of crime against property from 2014 to 2023.

@@

# Discussion
The discussion section contains some speculation for the possible causes of the observed results, as well as some weaknesses and next steps of this study.

## Trends in Crime and the Socio-Economic Influence
From 2014 to 2023, Toronto experienced significant changes in crime rates, reflecting broader socio-economic trends and specific local conditions. The steady increase in total crime, with a substantial rise from 113,426 cases in 2014 to 169,620 in 2023, partly reflects population growth and the increased digitalization of society. Crimes like theft and property-related offenses are strongly influenced by socio-economic conditions, including unemployment rates and economic hardship, especially during events like the COVID-19 pandemic. Financial instability likely contributed to the rise in financially motivated crimes such as fraud, which saw a sharp increase (124.67% growth rate) during this period, as indicated in Table 6​.

## Impact of the COVID-19 Pandemic
The effects of the COVID-19 pandemic are reflected in the 2020 and 2021 data. The lockdowns, which reduced public interactions, led to a noticeable decline in traditional crimes like burglaries but fueled an increase in cybercrime and domestic violence. Notably, the pandemic likely caused a shift in crime patterns, with the overall decrease in reported crimes in 2020 (118,568 cases) reflecting fewer opportunities for certain physical crimes​. However, fraud, particularly technology-facilitated fraud, continued to rise, highlighting the changing nature of crime in a digital era​.

## Crime Subtypes and Clear Rates
Despite the overall increase in reported crime, the clearing rate for most crime subtypes has declined substantially, with fraud experiencing a significant drop, from 38.55% in 2014 to just 4.99% in 2023. This drop might reflect the increasing complexity of investigating technology-facilitated crimes, where criminals are often harder to identify and prosecute​. The decrease in clear rates across all subtypes could suggest that law enforcement resources may struggle to keep pace with the rising volume and sophistication of crimes, particularly in property and financial crimes​.

## Weaknesses and next steps
One limitation of the analysis is the aggregation of crime data across various divisions, which might obscure regional differences within Toronto. Future research could benefit from a more granular analysis at the division level to capture localized crime trends and their underlying causes. Furthermore, while this report focuses on quantitative data, a qualitative approach examining the socio-economic drivers in more detail could provide deeper insights into the rise in specific crime subtypes, especially tech-related fraud.
Another significant weakness is that this dataset only includes reported crimes. Many crimes, such as domestic violence, may go unreported due to fear, shame, or a lack of trust in law enforcement. This underreporting skews the data and limits the ability to understand the prevalence of such crimes fully. A more comprehensive approach would require gathering data from additional sources, such as surveys on victimization, to capture the full scope of criminal activity in Toronto.
Moving forward, addressing the root causes of crime, particularly economic hardship and digital vulnerabilities, will be crucial in reducing crime rates and improving law enforcement's ability to clear cases. Enhanced training for officers handling digital crimes and investing in technologies to track and combat cybercrimes could improve clearance rates, particularly for fraud cases.
